---
title: "Codebook.Rmd"
author: "[Patrick (Pat) T...](https://github.com/pjtoms/ds3-gcd)"
output:
  md_document:
    variant: markdown_github
    toc: TRUE
---
#Codebook.md
## General Information
* __Author:__ *pjtx4657, Patrick (Pat) T.*
* __GitHub Project Link:__ [project](https://github.com/pjtoms/ds3-gcd "GitHub project link")
* __Last update:__ `r Sys.time()`<br>

## Study Design
Getting And Cleaning Data - Coursera Data Science Course 3, demonstrates ability to collect, work with, and clean a data set. The original data is manipulated into summarized by activity and subject [tidy](http://vita.had.co.nz/papers/tidy-data.pdf "tidy data paper") form.

The data linked to from the course website represents data collected from the accelerometers of the Samsung Galaxy S smartphone. A full description is available at the [site](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones "from UCI Machine Learning Repository - citation request at bottom of page") where the data was obtained. And the raw data for the project is located [here](https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip "from UCI Machine Learning Repository").

### Data Transformations
See [run_analysis.html](http://htmlpreview.github.com/?https://github.com/pjtoms/ds3-gcd/run_analysis.html "run_analysis.Rmd output") for a detailed description of data transformations, clean up, and the code used in production of the tidy data file.

### Definitions
More detail is available from the UCI raw data referenced above.

```{r terms, echo=FALSE, results='asis'}
cat("Term | Definition\n")
cat("------------------ |:----------------------------------\n")
cat("Domain | Category for signal derivation. Signals periodically obtained are part of the 'Time' domain. Signals processed using Fast Fourier Transform (FFT) are part of the 'Frequency' domain.\n")
cat("Accelerometer | Signal generator in 3-axial (X, Y, Z) directions.\n")
cat("Gyroscope | Signal generator in 3-axial (X, Y, Z) directions.\n")
cat("Body | Part of acceleration signal.\n")
cat("Gravity | Part of acceleration signal.\n")
cat("Jerk | Body linear acceleration and angular velocity derived in time domain.\n")
cat("Magnitude | Magnitude of the three-dimensional signals calculated using the Euclidean norm.\n")
cat("Signal | The actual telemetry received from the device.\n")
cat("Mean of Means | The data has been averaged for some values and extracted for this project. These values are then averaged by activity and subject.\n")
cat("Mean of Standard Deviations | The data has calculated standard deviations for the values which are extracted for this project. These values are then averaged by activity and subject.\n")
```

**Note:** There may be NA values in the tidy data due to factor usage or other missing values.

### Some R setup for producing the Code Book.
Code is hidden. See Codebook.Rmd for details.
```{r setup, echo=FALSE}

# Get full name descriptions for data.
namesDict <- "./namesDictionary.txt"
dtNamesDict <- data.table(read.table(namesDict, header=TRUE))
setkey(dtNamesDict, "Variable")

# Get regular expression based descriptions for data.
# Don't set a key, the table order should be used ASIS.
grepDict <- "./grepDictionary.txt"
dtGrepDict <- data.table(read.table(grepDict, header=TRUE))

getDescAndType <- function(name) {
  columnDescription <- ""
  descAndType <- ""
  # Flag to denote we formed a codeboook entry.
  matchFound <- FALSE
  
  if (as.logical(nrow(dtNamesDict[name, nomatch=0L]))) {
    descAndType <- cat(name, " | ", format(dtNamesDict[name]$Description), " | ", format(dtNamesDict[name]$Value), "\n")
    matchFound <- TRUE
  } else {
    for (j in 1:nrow(dtGrepDict)) {
      if (grepl(format(dtGrepDict[j]$Expression), name)) {
        matchFound <- TRUE
        columnDescription <- paste(columnDescription, format(dtGrepDict[j]$Text))
        if (!as.logical(dtGrepDict[j]$Continue)) {
          break
        }
      }
    }
    
    if (matchFound) {
      descAndType <- cat(name, " | ", columnDescription, " | Number between -1 and 1.\n")
    }
  }
  
  if (!matchFound) {
    descAndType <- cat(name, " | unknown desc | unknown type \n")
  }
}

```

## Code Book

Generated based on column names and contents of "namesDictionary.txt" and "grepDictionary.txt".

Code is hidden. See Codebook.Rmd for details.

```{r echo=FALSE, results='asis'}
cat("Variables | Descriptions | Value Type\n")
cat("---------------------- |:----------------------|----------------------\n")
for (i in namesVector) {
  cat(getDescAndType(i))
}
```

## Additional Information

### Tidy Data Structure
```{r echo=FALSE}
str(dtTidy)
```

### To Use the Tidy Data in R
The tidy data is saved as an R data.table in text and CSV formats. The following sample code may be used to read the data.table and restore its types as needed.

I don't run this code, just provide it as an example.

```{r eval=FALSE, echo=TRUE}

#
# To input the tidy 'wide' format data (my default preference) as an R data.table.
#
dtTidy <- read.table(tidy_txt_fname, stringsAsFactors = FALSE, header=TRUE)
dtTidy[, c(1)] <- sapply(dtTidy[, c(1)], as.factor)
# dtTidy is now in the same format as what was saved.

#
# To input the tidy 'tall' format data as an R data.table.
#
dtTidy <- read.table(tidy_txt_fname, stringsAsFactors = FALSE, header=TRUE)

# I think the following should work but doesn't on my system.
dtTidy[, c(1, 3:8)] <- sapply(dtTidy[, c(1, 3:8)], as.factor)

# This gets me back to the original form.
dtTidy[, c(1)] <- sapply(dtTidy[, c(1)], as.factor)
# And the following repeated, replacing 'x' with the number 3 through 8.
dtTidy[, c(x)] <- sapply(dtTidy[, c(x)], as.factor)
# dtTidy is now in the same format as what was saved.

```
